---
title: "Owl_Monkey_Mass"
output:
  html_document:
    toc: true 
    toc_float: true
---

Warning: this code will take up to an hour to run all the way through. 
You can find code alongside its output in the knitted html document.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = TRUE)
```

### Packages
This includes the packages used in the code.
```{r packages}
library(dplyr)
library(tidyr)
library(ggplot2)
library(lubridate)
library(brms)
library(ggdag)
library(pals)
library(cetcolor)
library(lme4)
library(ggtext)
library(parallel)
library(truncnorm)
library(purrr)
library(MASS)
library(zoo)
library(Hmisc)
```

# Temperature

## Temperature Data
This is the hourly temperature data from El Pucú Airport in Formosa (hourly). 
It has also been summarized at the day-level scale (daily).
```{r temperaturepucu, cache=TRUE}
# load hourly temperature data and separate date and time components
hourly <- 
  read.csv("./Data_and_Metadata/ElPucuAirportTemperature/Pucuhourlytemp.csv") |>
  separate(DATE, into = c("Date", "Time"), sep = "T")

# convert date column to date format
hourly$Date <- as.Date(hourly$Date)

# filter to retain only hourly records at the start of each hour
hourly <- hourly |>
  filter(substr(Time, 4, 5) == "00" & substr(Time, 7, 8) == "00")

# aggregate hourly data into daily summaries
daily <- hourly %>%
  group_by(Date) %>%
  filter(sum(!is.na(temp)) / n() >= 0.95) %>%  # retain days with at least 95% valid data
  dplyr::summarize(
    mean_temp = mean(temp, na.rm = TRUE),  # calculate mean temperature
    prop_above_35 = mean(temp > 35, na.rm = TRUE),  # proportion of hourly temperatures above 35
    prop_below_20 = mean(temp < 20, na.rm = TRUE)  # proportion of hourly temperatures below 20
  )

# add a column for month and day
daily <- mutate(daily, month_day = format(Date, "%m-%d"))

# calculate the day of the year from month and day
daily$day_of_year <- as.numeric(format(as.Date(daily$month_day, format = "%m-%d"), "%j"))

# process hourly data with similar columns
hourly <- hourly %>%
  mutate(
    month_day = format(Date, "%m-%d"),
    day_of_year = as.numeric(format(as.Date(month_day, format = "%m-%d"), "%j"))
  )

```

This is the temperature data from Estancia Guaycolec (temperature_eg). 
It has also been summarized at the day-level scale (temperature_eg_daily).
```{r temperatureeg, cache=TRUE}
# read temperature data and convert date column
temperature_eg <- 
  read.csv("./Data_and_Metadata/EstanciaGuaycolecTemperature/EstanciaGuaycolecTemp.csv")
temperature_eg$Date <- as.Date(temperature_eg$Date)

# create daily summaries for temperature data
temperature_eg_daily <- temperature_eg %>%
  group_by(Date) %>%
  filter(sum(!is.na(Temperature)) / n() >= 0.95) %>%  # retain days with at least 95% valid data
  dplyr::summarize(
    mean_temp = mean(Temperature, na.rm = TRUE),  # mean temperature
    prop_above_35 = mean(Temperature > 35, na.rm = TRUE),  # proportion above 35
    prop_below_20 = mean(Temperature < 20, na.rm = TRUE),  # proportion below 20
    variance = var(Temperature, na.rm = TRUE)  # variance of temperature
  )

# add columns for month-day and day of the year to daily data
temperature_eg_daily <- temperature_eg_daily %>%
  mutate(
    month_day = format(Date, "%m-%d"),
    day_of_year = as.numeric(format(as.Date(month_day, format = "%m-%d"), "%j"))
  )

# repeat similar processing for the full dataset
temperature_eg <- temperature_eg %>%
  mutate(
    month_day = format(Date, "%m-%d"),
    day_of_year = as.numeric(format(as.Date(month_day, format = "%m-%d"), "%j"))
  )
```

## Temperature Functions
These functions calculate temperature values in the past month. 
```{r lastmonthfunctions, cache=TRUE}
last_month_temp <- function(Date) {
  # convert input to a date object
  input_date <- as.Date(Date)
  
  # create a range of dates covering the last 30 days
  date_range <- input_date - 30:0
  
  # subset the data for the specified date range
  subset_data <- filter(daily, Date %in% date_range)
  
  # check if at least 95% of the data is available
  if (length(subset_data$mean_temp) / 31 >= 0.95) {
    # calculate the average temperature
    avg_temp <- mean(subset_data$mean_temp, na.rm = TRUE)
    return(avg_temp)
  } else {
    return(NA)  # return NA if insufficient data
  }
}

last_month_temp_ut_prop <- function(Date) {
  # convert input to a date object
  input_date <- as.Date(Date)
  
  # create a range of dates covering the last 30 days
  date_range <- input_date - 30:0
  
  # subset the data for the specified date range
  subset_data <- filter(daily, Date %in% date_range)
  
  # check if at least 95% of the data is available
  if (length(subset_data$prop_above_35) / 31 >= 0.95) {
    # calculate the mean proportion above 35
    mean_above_35 <- mean(subset_data$prop_above_35, na.rm = TRUE)
    return(mean_above_35)
  } else {
    return(NA)  # return NA if insufficient data
  }
}
```

These functions calculate temperature values in the past year. 
```{r lastyearfunctions, cache=TRUE}
last_year_temp <- function(Date) {
  # convert input to a date object
  input_date <- as.Date(Date)
  
  # create a range of dates covering the last 365 days
  date_range <- input_date - 365:0
  
  # subset the data for the specified date range
  subset_data <- filter(daily, Date %in% date_range)
  
  # check if at least 95% of the data is available
  if (length(subset_data$mean_temp) / 365 >= 0.95) {
    # calculate the average temperature
    avg_temp <- mean(subset_data$mean_temp, na.rm = TRUE)
    return(avg_temp)
  } else {
    return(NA)  # return NA if insufficient data
  }
}

last_year_temp_ut_prop <- function(Date) {
  # convert input to a date object
  input_date <- as.Date(Date)
  
  # create a range of dates covering the last 365 days
  date_range <- input_date - 365:0
  
  # subset the data for the specified date range
  subset_data <- filter(daily, Date %in% date_range)
  
  # check if at least 95% of the data is available
  if (length(subset_data$prop_above_35) / 365 >= 0.95) {
    # calculate the mean proportion above 35
    mean_above_35 <- mean(subset_data$prop_above_35, na.rm = TRUE)
    return(mean_above_35)
  } else {
    return(NA)  # return NA if insufficient data
  }
}

last_year_temp_lt_prop <- function(Date) {
  # convert input to a date object
  input_date <- as.Date(Date)
  
  # create a range of dates covering the last 365 days
  date_range <- input_date - 365:0
  
  # subset the data for the specified date range
  subset_data <- filter(daily, Date %in% date_range)
  
  # check if at least 95% of the data is available
  if (length(subset_data$prop_below_20) / 365 >= 0.95) {
    # calculate the mean proportion below 20
    mean_below_20 <- mean(subset_data$prop_below_20, na.rm = TRUE)
    return(mean_below_20)
  } else {
    return(NA)  # return NA if insufficient data
  }
}

```


# Dataset

```{r dataset, cache=TRUE}
#load data
biometrics7 <- read.csv("./Data_and_Metadata/FinalDataset/biometrics7.csv")

biometrics7$Sex <- as.factor(biometrics7$Sex)  # convert sex to factor
sexcolor_palette <- c("F" = "#498467", "M" = "#9067c6")  # color palette for sex

biometrics7$life_history <- factor(biometrics7$life_history, levels = c("predispersal", "floater", "territory_holder"))  # reorder life_history factor
lhcolor_palette <- c("predispersal" = "#DF6873", "floater" = "#AA5EE8", "territory_holder" = "#9DACFF")  # color palette for life history

# create ordered factors for birth status variables
biometrics7$BirthWithinYearfactor <- factor(biometrics7$BirthWithinYear, 
                                            levels = c(0, 2, 1), 
                                            labels = c("no birth", "unknown", "birth"), 
                                            ordered = TRUE)
biometrics7$BirthWithinNextYearfactor <- factor(biometrics7$BirthWithinNextYear, 
                                                levels = c(0, 2, 1), 
                                                labels = c("no birth", "unknown", "birth"), 
                                                ordered = TRUE)
biometrics7$BirthWithinLast5Monthsfactor <- factor(biometrics7$BirthWithinLast5Months, 
                                                   levels = c(0, 2, 1), 
                                                   labels = c("no birth", "unknown", "birth"), 
                                                   ordered = TRUE)

# standardize biometric data, and then unscale the variables I do not want to be scaled
numeric_columns <- sapply(biometrics7, is.numeric)
biometrics7_scaled <- biometrics7 |> 
  mutate(across(all_of(names(numeric_columns)[numeric_columns]), scale)) |> 
  mutate(day_of_year = biometrics7$day_of_year, 
         time_since_first_entry = biometrics7$time_since_first_entry,
         decades_since_first_entry = biometrics7$decades_since_first_entry,
         BirthWithinYear = biometrics7$BirthWithinYear,
         BirthWithinNextYear = biometrics7$BirthWithinNextYear,
         BirthWithinLast5Months = biometrics7$BirthWithinLast5Months,
         age = biometrics7$age)

# scale territory_holder data,
#and then unscale the variables I do not want to be scaled
biometrics7_scaled_th <- filter(biometrics7, life_history == "territory_holder")
biometrics7_scaled_th[, numeric_columns] <- scale(biometrics7_scaled_th[, numeric_columns])
biometrics7_scaled_th <- biometrics7_scaled_th |> mutate(day_of_year = filter(biometrics7, life_history == "territory_holder")$day_of_year) |>
  mutate(time_since_first_entry = filter(biometrics7, life_history == "territory_holder")$time_since_first_entry) |>
  mutate(decades_since_first_entry = filter(biometrics7, life_history == "territory_holder")$decades_since_first_entry) |>
  mutate(BirthWithinYear = filter(biometrics7, life_history == "territory_holder")$BirthWithinYear) |>
  mutate(BirthWithinNextYear = filter(biometrics7, life_history == "territory_holder")$BirthWithinNextYear) |>
  mutate(BirthWithinLast5Months = filter(biometrics7, life_history == "territory_holder")$BirthWithinLast5Months)

biometrics7$Date <- as.Date(biometrics7$Date)
biometrics7_scaled$Date <- as.Date(biometrics7_scaled$Date)
biometrics7_scaled_th$Date <- as.Date(biometrics7_scaled_th$Date)
```

# Summary statistics

## Table S1
```{r tables1, cache=TRUE,}
TableS1 <- biometrics7 |> 
  filter(!is.na(life_history)) |> 
  filter(!is.na(Weight)) |> 
  group_by(life_history, Sex) |> 
  dplyr::summarise(
    mean = round(mean(Weight, na.rm = T), -1),
    sd = round(sd(Weight, na.rm = T), -1),
    individuals = n_distinct(Subject),
    measurements = sum(!is.na(Weight))
  )

knitr::kable(TableS1)
```

## Body length summary stats
```{r bodylengthsamplesize}
bodylengthsamplesize <- biometrics7 |> 
  filter(!is.na(life_history)) |> 
  filter(!is.na(Bodylength)) |> 
  group_by(life_history, Sex) |> 
  dplyr::summarise(
    mean = round(mean(Bodylength, na.rm = T), 1),
    sd = round(sd(Bodylength, na.rm = T), 1),
    individuals = n_distinct(Subject),
    measurements = sum(!is.na(Bodylength))
  )

knitr::kable(bodylengthsamplesize)
```


## Effect sizes for models
```{r effectsizes, cache=TRUE}
th_only <- biometrics7 |> filter(life_history == "territory_holder") |>
  filter(!is.na(Weight)) |> 
   dplyr::summarize(
    mean = round(mean(Weight, na.rm = T), -1),
    sd = round(sd(Weight, na.rm = T), -1),
    individuals = n_distinct(Subject),
    measurements = sum(!is.na(Weight))
  )
knitr::kable(th_only)

all_lh <- biometrics7 |>
  filter(!is.na(Weight)) |> 
  filter(!is.na(life_history)) |> 
  dplyr::summarize(
    mean = round(mean(Weight, na.rm = T), -1),
    sd = round(sd(Weight, na.rm = T), -1),
    individuals = n_distinct(Subject),
    measurements = sum(!is.na(Weight))
  )

knitr::kable(all_lh)

all_lh_bodylength <- biometrics7 |>
  filter(!is.na(Bodylength)) |> 
  filter(!is.na(life_history)) |> 
  dplyr::summarize(
    mean = round(mean(Bodylength, na.rm = T), 1),
    sd = round(sd(Bodylength, na.rm = T), 1),
    individuals = n_distinct(Subject),
    measurements = sum(!is.na(Bodylength))
  )

knitr::kable(all_lh_bodylength)
```


# Bayesian Models

## Model 1
```{r model1, cache=TRUE}
# fit a bayesian regression model for weight with a cyclic spline for day_of_year,
# fixed effects for decades_since_first_entry and life_history, 
# and a random intercept for Subject
Model1 <- brm(
  formula = Weight ~ s(day_of_year, bs = "cc") + decades_since_first_entry + 
    life_history + (1 | Subject),  # model formula
  data = biometrics7_scaled,  # scaled dataset
  family = gaussian(),  # gaussian family for continuous outcome
  iter = 5000,  # number of iterations
  chains = 4,  # number of mcmc chains
  control = list(adapt_delta = 0.99),  # control settings to improve convergence
  prior = c(
    prior(normal(0, 0.7), class = "b", coef = "decades_since_first_entry"),  # prior for decades_since_first_entry
    prior(normal(1, 1), class = "b", coef = "life_historyfloater"),  # prior for life_history = floater
    prior(normal(1, 1), class = "b", coef = "life_historyterritory_holder"),  # prior for life_history = territory_holder
    prior(cauchy(0, 1), class = "sd", group = "Subject")  # prior for subject-level random effects
  ) 
)

summary(Model1)

# summarize the posterior results with credible intervals at 5.5% and 94.5%
summary_results1 <- posterior_summary(Model1, probs = c(0.055, 0.945), robust = TRUE)

# round the summary results to 2 decimal places for better readability
rounded_summary1 <- round(summary_results1, 2)

# view the rounded summary results
knitr::kable(rounded_summary1)

```
To interpret this, an effect size of one is equivalent to 170g.
2023 is 2.4 decades after 1999. An increase of 2.4 decades converts to:

```{r effectsizecalculation, cache=TRUE}
#calculation for effect size of decade:
2.4*170*0.28
#decade change * one standard deviation * effect size from Model 1 for decade

#should be about 110 g
```

an increase of ~110 grams, holding all other variables constant.


## Model 2
### Model 2F
```{r model2f, cache=TRUE}
# fit model 2 for females with weight as the outcome and predictors for 
#temperature, fruit, and birth factors
Model2F <- brm(
  formula = Weight ~ temp_new_year + twelvemonthfruit + BirthWithinYearfactor +
    s(day_of_year, bs = "cc") + 
    # include smooth term for day of year with cyclic cubic splines
    (1 | Subject),  # random effect for subject
  data = filter(biometrics7_scaled_th, Sex == "F"),  # filter data for females
  family = gaussian(),  # use gaussian distribution for the outcome
  iter = 5000,  # set number of iterations
  chains = 4,  # set number of chains for sampling
  control = list(adapt_delta = 0.999),  # increase adapt_delta for better convergence
  prior = c(
    prior(normal(0, 0.7), class = "b"),  # set normal prior for fixed effects
    prior(cauchy(0, 0.7), class = "sd", group = "Subject") 
    # set cauchy prior for random effects
  )
)

# summarize the posterior results for model 2f
summary_results2F <- posterior_summary(Model2F, probs = c(0.055, 0.945), robust = TRUE)
rounded_summary2F <- round(summary_results2F, 2)  # round summary results
knitr::kable(rounded_summary2F)
```

### Model 2M
```{r model2m, cache=TRUE}
# fit model 2 for males with weight as the outcome and predictors 
#for temperature #fruit, and birth factors
Model2M <- brm(
  formula = Weight ~ temp_new_year + twelvemonthfruit + BirthWithinYearfactor +
    s(day_of_year, bs = "cc") + 
    # include smooth term for day of year with cyclic cubic splines
    (1 | Subject),  # random effect for subject
  data = filter(biometrics7_scaled_th, Sex == "M"),  # filter data for males
  family = gaussian(),  # use gaussian distribution for the outcome
  iter = 5000,  # set number of iterations
  chains = 4,  # set number of chains for sampling
  control = list(adapt_delta = 0.999),  # increase adapt_delta for better convergence
  prior = c(
    prior(normal(0, 0.7), class = "b"),  # set normal prior for fixed effects
    prior(cauchy(0, 0.7), class = "sd", group = "Subject") 
    # set cauchy prior for random effects
  )
)

# summarize the posterior results for model 2m
summary_results2M <- posterior_summary(Model2M, probs = c(0.055, 0.945), robust = TRUE)
rounded_summary2M <- round(summary_results2M, 2)  # round summary results
knitr::kable(rounded_summary2M)

```

## Model 3
### Model 3F
```{r model3f, cache=TRUE}
# fit the model for females (Model3F)
Model3F <- brm(
  formula = Weight ~  temp_new + onemonthfruit + BirthWithinLast5Monthsfactor +
    decades_since_first_entry + (1 | Subject), 
  data = filter(biometrics7_scaled_th, Sex == "F"),
  family = gaussian(),
  iter = 5000, 
  chains = 4,
  control = list(adapt_delta = 0.999),
  prior = c(
    prior(normal(0, 0.7), class = "b"), 
    prior(cauchy(0, 0.7), class = "sd", group = "Subject")
  ))


summary_results3F <- posterior_summary(Model3F, 
                                       probs = c(0.055, 0.945), 
                                       robust = TRUE)
rounded_summary3F <- round(summary_results3F, 2)
knitr::kable(rounded_summary3F)
```

### Model 3M
```{r model3m, cache=TRUE}
# fit the model for males (Model3M)
Model3M <- brm(
  formula = Weight ~  temp_new + onemonthfruit + BirthWithinLast5Monthsfactor +
    decades_since_first_entry + (1 | Subject), 
  data = filter(biometrics7_scaled_th, Sex == "M"),
  family = gaussian(),
  iter = 5000, 
  chains = 4,
  control = list(adapt_delta = 0.999),
  # cores = parallel::detectCores(), 
  prior = c(
    prior(normal(0, 0.7), class = "b"), 
    prior(cauchy(0, 0.7), class = "sd", group = "Subject")
  ))

summary_results3M <- posterior_summary(Model3M, probs = c(0.055, 0.945), robust = TRUE)
rounded_summary3M <- round(summary_results3M, 2)
knitr::kable(rounded_summary3M)

```

## Model 4
```{r model4, cache=TRUE}
set.seed(13)

Model4 <- brm(
  formula = Weight ~ firstyearfruit + firstyeartemp_new + log(age, base = 2) + 
    (1 | Subject), 
  data = filter(biometrics7_scaled, age > 1),
  family = gaussian(),
  iter = 5000,
  chains = 4,
  prior = c(
    prior(normal(1, 2), class = "b"), 
    prior(normal(0, 1), class = "b", coef = "firstyearfruit"), 
    prior(normal(0, 1), class = "b", coef = "firstyeartemp_new"),
    prior(cauchy(0, 1), class = "sd", group = "Subject")
  ),
  # set control parameters for model fitting
  control = list(adapt_delta = 0.99)
)

summary_results4 <- posterior_summary(Model4, probs = c(0.055, 0.945), robust = TRUE)
rounded_summary4 <- round(summary_results4, 2)
knitr::kable(rounded_summary4)
```

## Model 5
```{r model5, cache=TRUE}
Model5 <- brm(
  formula = Bodylength ~ s(day_of_year, bs = "cc") + decades_since_first_entry + 
    life_history + (1 | Subject), 
  data = biometrics7_scaled,
  family = gaussian(),
  iter = 6000, 
  chains = 4,
  control = list(adapt_delta = 0.99),
  prior = c(
    prior(normal(0, 0.7), class = "b", coef = "decades_since_first_entry"),
   prior(normal(1, 1), class = "b", coef = "life_historyfloater"),
    prior(normal(1, 1), class = "b", coef = "life_historyterritory_holder"),
    prior(cauchy(0, 1), class = "sd", group = "Subject")
  ),
  cores = parallel::detectCores()
)

# Summarize the model
summary(Model5) 

summary_resultsModel5 <- 
  posterior_summary(Model5, probs = c(0.055, 0.945), robust = TRUE)
rounded_summaryModel5 <- round(summary_resultsModel5, 2)
knitr::kable(rounded_summaryModel5)
```

# Simulation
```{r simulation, cache=TRUE}
set.seed(13)
th_narm <- biometrics7 |> filter(life_history == "territory_holder") |>
  filter(!is.na(Weight)) |> filter(BirthWithinYearfactor != "unknown")

# number of bootstrap iterations
n_iterations <- 20000

# function to perform a single bootstrap iteration
bootstrap_function <- function(data) {
  # resample the data with replacement
  bootstrap_sample <- data |>
    sample_n(nrow(data), replace = TRUE)
  
  # calculate mean weight for each BirthWithinYearfactor in the bootstrap sample
  means <- bootstrap_sample |>
    group_by(BirthWithinYearfactor) |>
    dplyr::summarize(mean_weight = mean(Weight, na.rm = TRUE))
  
  # extract means for "birth" and "no birth"
  mean_birth <- means |>
    filter(BirthWithinYearfactor == "birth") |>
    pull(mean_weight)
  mean_no_birth <- means |>
    filter(BirthWithinYearfactor == "no birth") |>
    pull(mean_weight)
  
  # calculate the difference between means
  diff <- mean_birth - mean_no_birth
  
  # generate a random sample from a truncated normal distribution (0, 1)
  random_value <- rtruncnorm(1, a = 0, b = 1, mean = 0.7, sd = 0.2)
  
  # calculate the product
  product <- diff * random_value
  
  # return a data frame with the results
  return(tibble(bootstrapped_diff = diff, random_value = random_value, product = product))
}

# repeat the bootstrap process
bootstrap_results <- map_df(1:n_iterations, ~bootstrap_function(th_narm))
```

### Simulation results
```{r simulationresults, cache=TRUE}
# calculate mean and 89% interval quantiles for bootstrapped_diff
bootstrapped_diff_mean <- mean(bootstrap_results$bootstrapped_diff, na.rm = TRUE)
bootstrapped_diff_lower <- quantile(bootstrap_results$bootstrapped_diff, 0.055, na.rm = TRUE)
bootstrapped_diff_upper <- quantile(bootstrap_results$bootstrapped_diff, 0.945, na.rm = TRUE)

# calculate mean and 89% interval quantiles for random_value
random_value_mean <- mean(bootstrap_results$random_value, na.rm = TRUE)
random_value_lower <- quantile(bootstrap_results$random_value, 0.055, na.rm = TRUE)
random_value_upper <- quantile(bootstrap_results$random_value, 0.945, na.rm = TRUE)

# calculate mean and 89% interval quantiles for product
product_mean <- mean(bootstrap_results$product, na.rm = TRUE)
product_lower <- quantile(bootstrap_results$product, 0.055, na.rm = TRUE)
product_upper <- quantile(bootstrap_results$product, 0.945, na.rm = TRUE)

# output results in a readable, formatted text block
cat(paste("Selective differential:\n",
          "  Mean: ", round(bootstrapped_diff_mean, 3), "\n",
          "  89% Interval: [", round(bootstrapped_diff_lower, 3), ", ", round(bootstrapped_diff_upper, 3), "]\n\n",
          
          "Heritability:\n",
          "  Mean: ", round(random_value_mean, 3), "\n",
          "  89% Interval: [", round(random_value_lower, 3), ", ", round(random_value_upper, 3), "]\n\n",
          
          "Response to selection:\n",
          "  Mean: ", round(product_mean, 3), "\n",
          "  89% Interval: [", round(product_lower, 3), ", ", round(product_upper, 3), "]\n", sep = ""))
```


# Visualizations
## Figure 1
```{r figure1, cache=TRUE}
ggplot(filter(biometrics7, life_history %in% c("territory_holder", "floater", "predispersal")), aes(x = Date, y = Weight)) +
  # scatter plot with color and shape by life history
  geom_point(aes(color = life_history, shape = life_history), size = 4, alpha = 0.7) +
  # manually set color palette for life history
  scale_color_manual(values = lhcolor_palette) +
  # add linear model smooth line with shaded confidence interval
  geom_smooth(method = "lm", color = "black", se = TRUE, fill = "lightgray", level = 0.89) +
  # facet by sex and life history with custom labels
  facet_grid(Sex ~ life_history,
             labeller = labeller(
               Sex = c("M" = "Male", "F" = "Female"),
               life_history = c("predispersal" = "Predispersal",
                                "floater" = "Floater",
                                "territory_holder" = "Reproductive adult")
             )) +
  # minimal theme for cleaner look
  theme_minimal() +
  # customize axis labels and theme elements
  labs(x = "Year", y = "Mass\n(g)") +
  theme(
    text = element_text(family = "Helvetica", size = 14),
    axis.text = element_text(size = 14),
    axis.text.y = element_text(size = 12, angle = 0, vjust = 0.5),
    axis.text.x = element_text(size = 9),
    axis.title.y = element_text(angle = 0, vjust = 0.5, size = 14),
    panel.spacing = unit(1, "lines"),
    strip.text.x = element_text(size = 12),  # styling for life history labels
    strip.text.y = element_text(size = 12, angle = 0, hjust = 0)  # styling for sex labels
  ) +
  # remove the legend for color and shape
  guides(color = "none", shape = "none") +
  # customize x-axis with date limits and breaks
  scale_x_date(
    limits = as.Date(c("1999-01-01", "2023-12-31")),
    breaks = seq(as.Date("1999-01-01"), as.Date("2023-12-31"), by = "6 years"),
    date_labels = "%Y"
  )

```

## Figure 2
```{r figure2, cache=TRUE}
# set window size (365 days) for rolling calculations
window_size <- 365

# calculate rolling mean temperature and coverage for the daily dataset
daily <- daily %>%
  arrange(Date) %>%  # arrange the data by Date
  mutate(
    # calculate rolling mean temperature using a window of 365 days
    mean_temp_roll = rollapply(mean_temp, width = window_size, FUN = mean, fill = NA, align = "center"),
    
    # calculate coverage as the fraction of unique days in the window
    coverage = rollapply(Date, width = window_size, FUN = function(dates) {
      # get the range of dates in the window
      date_range <- seq(min(dates), max(dates), by = "1 day")
      
      # calculate the fraction of unique days in the date range
      unique_days <- length(unique(dates))
      total_days <- length(date_range)
      
      # return coverage as a percentage
      return((unique_days / total_days) * 100)
    }, fill = NA, align = "center")
  )

# filter daily dataset for mean_temp_roll where coverage is at least 95%
filtered_daily <- daily %>% filter(coverage >= 95)

# plot mean temperature for filtered data with a smooth line
ggplot(filtered_daily, aes(x = Date, y = mean_temp_roll)) +
  # scatter plot of mean temperature with black points
  geom_point(color = "black", size = 0.4, alpha = 0.75) + 
  
  # add a smoothed line using loess method with a maroon color and no confidence interval
  geom_smooth(span = 0.4, method = "loess", color = "maroon", linewidth = 1.5, se = F) + 
  
  # apply minimal theme
  theme_minimal() +
  
  # add labels for y and x axes
  labs(y = "365-day\nmean\ntemperature\n(ºC)", x = "Year") +
  
  # customize x-axis with specific limits and breaks for every 6 years
  scale_x_date(
    limits = as.Date(c("1998-01-01", "2023-12-31")),
    breaks = seq(as.Date("1999-01-01"), as.Date("2024-12-31"), by = "6 years"),
    date_labels = "%Y"
  ) +
  
  # customize text size and appearance for axis labels and titles
  theme(
    text = element_text(family = "Helvetica", size = 15),
    axis.text = element_text(size = 14),
    axis.text.y = element_text(size = 12, angle = 0, vjust = 0.5),
    axis.text.x = element_text(size = 10),
    axis.title.y = element_text(angle = 0, vjust = 0.5, size = 13)
  )

```

## Figure 3
This is slightly more complicated, requiring using Model 4 to adjust appropriately.
```{r figure3, cache=TRUE}
# step 1: calculate the predicted values without firstyeartemp_new
# extract predictions from the model with all terms except firstyeartemp_new
fitted_no_temp <- predict(Model4, allow_new_levels = T, newdata = biometrics7_scaled |>
                           filter(age > 1) |>
                           mutate(firstyeartemp_new = 0))[, "Estimate"]

# step 2: subtract these predicted values from the observed data
biometrics7_scaled_subtracted <- biometrics7_scaled |>
  filter(age > 1) |>
  mutate(
    adjusted_weight = Weight - fitted_no_temp  # adjusted weight after controlling for other variables
  )


# Generate conditional effects with an 89% credible interval
temp_effects <- conditional_effects(Model4, effects = "firstyeartemp_new", probs = c(0.055, 0.945))

# Extract the conditional effects data
temp_data <- temp_effects$firstyeartemp_new

# step 3: plot the adjusted points along with the conditional effects
ggplot() +
  # plot the conditional effects of temperature
  geom_ribbon(data = temp_data, aes(x = firstyeartemp_new, ymin = lower__, ymax = upper__), alpha = 0.2, fill = "lightgray") +
  
  # plot the adjusted points with custom shapes for life_history
  geom_point(data = biometrics7_scaled_subtracted, 
             aes(x = firstyeartemp_new, y = adjusted_weight, color = life_history, shape = life_history), 
             alpha = 0.7, size = 6) +
  
  # customize shapes: square for "territory_holder", circle for "predispersal"
  scale_shape_manual(values = c("territory_holder" = 15, "predispersal" = 16)) + 
  scale_color_manual(values = lhcolor_palette) +
  
  # add the prediction line
  geom_line(data = temp_data, aes(x = firstyeartemp_new, y = estimate__), color = "black") +
  
  # customize labels
  labs(
    x = "mean temperature during\nfirst year of life (standardized)",
    y = "body mass\n(adjusted and\nstandardized)"
  ) + 
  # make y-axis label horizontal and centered
  theme_minimal() + 
  theme(text = element_text(family = "Helvetica", size = 14),
        axis.text = element_text(size = 14),
        axis.text.y = element_text(size = 12, angle = 0, vjust = 0.5),
        axis.text.x = element_text(size = 9),
        axis.title.y = element_text(angle = 0, vjust = 0.5, size = 14)) + 
  guides(color = FALSE, shape = FALSE) + 
  xlim(NA, 1.5)
```

## Figure 4
```{r figure4, cache=TRUE}
# bin the bootstrapped diff and random value with reduced number of bins
bootstrap_results_binned <- bootstrap_results |>
  mutate(
    bootstrapped_diff_bin = cut(bootstrapped_diff, breaks = seq(-75, 75, by = 25)),  # reduced number of bins
    random_value_bin = cut(random_value, breaks = seq(0, 1, by = 0.2))  # reduced number of bins
  ) |>
  group_by(bootstrapped_diff_bin, random_value_bin) |>
  dplyr::summarize(
    mean_product = mean(product, na.rm = TRUE),  # calculate mean of product for each bin
    .groups = 'drop'  # avoid keeping grouping structure
  )

# create the heatmap with improved axis readability
ggplot(filter(bootstrap_results_binned, !is.na(bootstrapped_diff_bin)), aes(x = bootstrapped_diff_bin, y = random_value_bin, fill = mean_product)) +
  geom_tile() +  # create heatmap tiles
  scale_fill_gradient2(low = "red", mid = "#FAF9F6", high = "blue") +  # custom color gradient
  theme_minimal() +  # minimal theme for cleaner look
  theme(
    text = element_text(family = "Helvetica", size = 15),  # set font family and size for text
    axis.text.x = element_text(size = 12, angle = 30, hjust = 1),  # rotate x-axis labels for better readability
    axis.text.y = element_text(size = 12),  # increase size of y-axis labels
    axis.title = element_text(size = 14),  # set axis title size
    axis.title.y = element_text(angle = 0, vjust = 0.5, size = 14),  # vertical adjustment for y-axis title
    axis.ticks.x = element_blank(),  # remove x-axis ticks
    axis.ticks.y = element_blank()   # remove y-axis ticks
  ) + 
  labs(x = "\nSelective differential (g)", y = "Heritability   ", fill = "Response to\nselection (g)")  # customize axis labels

cat(paste("Selective differential:\n",
          "  Mean: ", round(bootstrapped_diff_mean, 3), "\n",
          "  89% Interval: [", round(bootstrapped_diff_lower, 3), ", ", round(bootstrapped_diff_upper, 3), "]\n\n",
          
          "Heritability:\n",
          "  Mean: ", round(random_value_mean, 3), "\n",
          "  89% Interval: [", round(random_value_lower, 3), ", ", round(random_value_upper, 3), "]\n\n",
          
          "Response to selection:\n",
          "  Mean: ", round(product_mean, 3), "\n",
          "  89% Interval: [", round(product_lower, 3), ", ", round(product_upper, 3), "]\n", sep = ""))
```

## Figure S1A
```{r figures1a, cache=TRUE}
# define von bertalanffy growth equation
von_bertalanffy <- function(age, A, K, t0) { 
  A * (1 - exp(-K * (age - t0))) 
}

# filter out NA values
filtered_data <- biometrics7 |> 
  filter(!is.na(life_history))

# fit von bertalanffy curve to data with different starting values and optimization algorithm
fit <- nls(Weight ~ von_bertalanffy(age, A, K, t0),  
           data = filtered_data,
           start = list(A = 1000, K = 0.1, t0 = 0),  # adjust starting values
           algorithm = "port")  # try a different optimization algorithm

# extract coefficients
coefficients <- coef(fit)

# plot data and von bertalanffy curve
ggplot(filtered_data, aes(x = age, y = Weight)) + 
  geom_point(size = 4, alpha = 0.7, aes(color = life_history, shape = life_history)) + 
  geom_smooth(method = "nls",  
              formula = y ~ von_bertalanffy(x, A, K, t0),  
              se = FALSE,  
              color = "black", 
              method.args = list(start = coefficients)) + 
  scale_color_manual( 
    values = lhcolor_palette,  
    labels = c("predispersal" = "Predispersal",  
               "floater" = "Floater",  
               "territory_holder" = "Reproductive adult") 
  ) + 
  scale_shape_manual( 
    values = c("predispersal" = "circle",  # you can adjust shapes as needed
               "floater" = "triangle",  
               "territory_holder" = "square"),  
    labels = c("predispersal" = "Predispersal",  
               "floater" = "Floater",  
               "territory_holder" = "Reproductive adult") 
  ) + 
  labs(x = "Age (years)", y = "Body\nmass\n(g)",  
       color = "Life History Stage", shape = "Life History Stage") + 
  theme_minimal() + 
  scale_x_continuous( 
    limits = c(0, 8), 
    breaks = seq(0, 8, by = 2) 
  ) + 
  theme( 
    text = element_text(family = "Helvetica", size = 15), 
    axis.text = element_text(size = 14), 
    axis.text.y = element_text(size = 12, angle = 0, vjust = 0.5), 
    axis.text.x = element_text(size = 12), 
    axis.title.y = element_text(angle = 0, vjust = 0.5, size = 14)
  )
```

## Figure S1B
```{r figures1b, cache=TRUE}
# define von bertalanffy growth equation
von_bertalanffy <- function(age, A, K, t0) { 
  A * (1 - exp(-K * (age - t0))) 
}

# filter out NA values and bodylength greater than 40
filtered_data <- biometrics7 |> 
  filter(!is.na(life_history)) |> 
  filter(Bodylength > 40)

# fit von bertalanffy curve to bodylength data with adjusted starting values and iteration limit
fit_bodylength <- nls(Bodylength ~ von_bertalanffy(age, A, K, t0),  
                      data = filtered_data,
                      start = list(A = 500, K = 0.05, t0 = 0),  # adjust starting values
                      algorithm = "port",  # use the "port" algorithm
                      control = list(maxiter = 100))  # increase iteration limit

# extract coefficients for bodylength
coefficients_bodylength <- coef(fit_bodylength)

# plot bodylength data and von bertalanffy curve
ggplot(filtered_data, aes(x = age, y = Bodylength)) + 
  geom_point(size = 4, alpha = 0.7, aes(color = life_history, shape = life_history)) + 
  geom_smooth(method = "nls",  
              formula = y ~ von_bertalanffy(x, A, K, t0),  
              se = FALSE,  
              color = "black", 
              method.args = list(start = coefficients_bodylength)) + 
  scale_color_manual( 
    values = lhcolor_palette,  
    labels = c("predispersal" = "Predispersal",  
               "floater" = "Floater",  
               "territory_holder" = "Reproductive adult") 
  ) + 
  scale_shape_manual( 
    values = c("predispersal" = "circle",  # you can adjust shapes as needed
               "floater" = "triangle",  
               "territory_holder" = "square"),  
    labels = c("predispersal" = "Predispersal",  
               "floater" = "Floater",  
               "territory_holder" = "Reproductive adult") 
  ) + 
  labs(x = "Age (years)", y = "Body\nlength\n(cm)",  
       color = "Life History Stage", shape = "Life History Stage") + 
  theme_minimal() + 
  scale_x_continuous( 
    limits = c(0, 8), 
    breaks = seq(0, 8, by = 2) 
  ) + 
  theme( 
    text = element_text(family = "Helvetica", size = 15), 
    axis.text = element_text(size = 14), 
    axis.text.y = element_text(size = 12, angle = 0, vjust = 0.5), 
    axis.text.x = element_text(size = 12), 
    axis.title.y = element_text(angle = 0, vjust = 0.5, size = 14)
  )
```

## Figure S2
```{r figures2, cache=TRUE}
# find the common dates between the two datasets
common_dates <- intersect(temperature_eg_daily$Date, daily$Date)

# filter each dataset to include only common dates
temperature_eg_daily_common <- filter(temperature_eg_daily, Date %in% as.Date(common_dates))
daily_common <- filter(daily, Date %in% as.Date(common_dates))

# plot the data with only common dates
ggplot() +
  # smooth line for temperature_eg_daily dataset
  geom_smooth(data = temperature_eg_daily_common, aes(x = day_of_year, y = mean_temp), color = "navy", 
              method = "gam", formula = y ~ s(x, bs = "cc")) + 
  # smooth line for daily dataset
  geom_smooth(data = daily_common, aes(x = day_of_year, y = mean_temp), color = "maroon", 
              method = "gam", formula = y ~ s(x, bs = "cc")) + 
  theme_minimal() + 
  labs(x = "Day of year", y = "Temperature\n(ºC)") + 
  theme(
    text = element_text(family = "Helvetica", size = 14),  # set text size and family
    axis.text = element_text(size = 14),  # set axis text size
    axis.text.y = element_text(angle = 0, vjust = 0.5, size = 12),  # adjust y-axis text size and alignment
    axis.text.x = element_text(size = 9),  # adjust x-axis text size
    axis.title.y = element_text(angle = 0, vjust = 0.5, size = 14),  # adjust y-axis title size
    panel.spacing = unit(1, "lines"),  # adjust panel spacing
    strip.text.x = element_text(size = 12),  # set label size for x-axis facets
    strip.text.y = element_text(size = 12, angle = 0, hjust = 0)  # set label size and alignment for y-axis facets
  )
```

## Figure S3
```{r figures3, cache=TRUE}
# define a function to calculate past 365-day coverage for a given dataset
calculate_coverage <- function(df) { 
  df |>
    arrange(Date) |>
    mutate(
      coverage = sapply(Date, function(current_date) {
        # define the past 365-day window
        start_date <- current_date - 364 
        # filter dates within this window
        date_window <- df$Date[df$Date >= start_date & df$Date <= current_date] 
        # calculate coverage as the fraction of unique days over the full 365 days
        coverage <- (length(unique(date_window)) / 365) * 100
        return(coverage)
      })
    )
}

# apply the coverage calculation function to each dataset
temperature_eg_daily <- calculate_coverage(temperature_eg_daily)
daily <- calculate_coverage(daily)

# plot the 365-day coverage lines for both datasets
ggplot() + 
  # 365-day coverage line for temperature_eg_daily
  geom_line(data = temperature_eg_daily, aes(x = Date, y = coverage), color = "navy") + 
  # 365-day coverage line for daily
  geom_line(data = daily, aes(x = Date, y = coverage), color = "maroon") + 
  # horizontal line at 95% coverage
  geom_hline(yintercept = 95, linetype = "dashed", color = "gray50") + 
  theme_minimal() + 
  labs(x = "Year", y = "365-day\ncoverage\n(%)") + 
  scale_y_continuous(limits = c(0, 100)) + 
  theme(
    text = element_text(family = "Helvetica", size = 14), # set text size and family
    axis.text = element_text(size = 14), # set axis text size
    axis.text.y = element_text(angle = 0, vjust = 0.5, size = 12), # adjust y-axis text size and alignment
    axis.text.x = element_text(size = 9), # adjust x-axis text size
    axis.title.y = element_text(angle = 0, vjust = 0.5, size = 14), # adjust y-axis title size
    panel.spacing = unit(1, "lines"), # adjust panel spacing
    strip.text.x = element_text(size = 12), # set label size for x-axis facets
    strip.text.y = element_text(size = 12, angle = 0, hjust = 0) # set label size and alignment for y-axis facets
  ) + 
  scale_x_date(
    limits = as.Date(c("1999-01-01", "2023-12-31")), 
    breaks = seq(as.Date("1999-01-01"), as.Date("2023-12-31"), by = "6 years"), 
    date_labels = "%Y"
  )
```

## Figure S4
```{r figures4, cache=TRUE}
# define a function to calculate past 30-day coverage for a given dataset
calculate_coverage_30_days <- function(df) { 
  df |>
    arrange(Date) |>
    mutate(
      coverage = sapply(Date, function(current_date) {
        # define the past 30-day window
        start_date <- current_date - 29 
        # filter dates within this window
        date_window <- df$Date[df$Date >= start_date & df$Date <= current_date] 
        # calculate coverage as the fraction of unique days over the full 30 days
        coverage <- (length(unique(date_window)) / 30) * 100
        return(coverage)
      })
    )
}

# apply the coverage calculation function to each dataset
temperature_eg_daily <- calculate_coverage_30_days(temperature_eg_daily)
daily <- calculate_coverage_30_days(daily)

# plot the 30-day coverage lines for both datasets
ggplot() + 
  # 30-day coverage line for temperature_eg_daily
  geom_line(data = temperature_eg_daily, aes(x = Date, y = coverage), color = "navy") + 
  # 30-day coverage line for daily
  geom_line(data = daily, aes(x = Date, y = coverage), color = "maroon") + 
  # horizontal line at 95% coverage
  geom_hline(yintercept = 95, linetype = "dashed", color = "gray50") + 
  theme_minimal() + 
  labs(x = "Year", y = "30-day\ncoverage\n(%)") + 
  scale_y_continuous(limits = c(0, 100)) + 
  theme(
    text = element_text(family = "Helvetica", size = 14), # set text size and family
    axis.text = element_text(size = 14), # set axis text size
    axis.text.y = element_text(angle = 0, vjust = 0.5, size = 12), # adjust y-axis text size and alignment
    axis.text.x = element_text(size = 9), # adjust x-axis text size
    axis.title.y = element_text(angle = 0, vjust = 0.5, size = 14), # adjust y-axis title size
    panel.spacing = unit(1, "lines"), # adjust panel spacing
    strip.text.x = element_text(size = 12), # set label size for x-axis facets
    strip.text.y = element_text(size = 12, angle = 0, hjust = 0) # set label size and alignment for y-axis facets
  ) + 
  scale_x_date(
    limits = as.Date(c("1999-01-01", "2023-12-31")), 
    breaks = seq(as.Date("1999-01-01"), as.Date("2023-12-31"), by = "6 years"), 
    date_labels = "%Y"
  )
```

## Figure S5A
```{r figures5a, cache=TRUE}
ggplot(filter(biometrics7, !is.na(life_history) & !is.na(Sex)), aes(x = temp_new_year, y = Weight)) + 
  geom_point(aes(color = life_history, shape = life_history), size = 4, alpha = 0.7) +
  scale_color_manual(values = lhcolor_palette) +
  geom_smooth(method = "lm", color = "black", se = TRUE, fill = "lightgray", level = 0.89) +
  facet_grid(Sex ~ life_history, 
             labeller = labeller(
               Sex = c("M" = "Male", "F" = "Female"),
               life_history = c("predispersal" = "Predispersal",
                                "floater" = "Floater",
                                "territory_holder" = "Reproductive adult")
             )) +
  theme_minimal() +
  labs(x = "Mean temperature in past year (ºC)", y = "Mass\n(g)") +
  theme(text = element_text(family = "Helvetica", size = 14),
        axis.text = element_text(size = 14),
        axis.text.y = element_text(size = 12, angle = 0, vjust = 0.5),
        axis.text.x = element_text(size = 12),
        axis.title.y = element_text(angle = 0, vjust = 0.5, size = 14),
        panel.spacing = unit(1.3, "lines"),
        strip.text.x = element_text(size = 9),  
        strip.text.y = element_text(size = 12, angle = 0, hjust = 0)) + 
  guides(color = FALSE, shape = FALSE) 
```

## Figure S5B
```{r figures5b, cache=TRUE}
ggplot(filter(biometrics7, !is.na(life_history) & !is.na(Sex)), aes(x = temp_ltprop_year, y = Weight)) + 
  geom_point(aes(color = life_history, shape = life_history), size = 4, alpha = 0.7) +
  scale_color_manual(values = lhcolor_palette) +
  geom_smooth(method = "lm", color = "black", se = TRUE, fill = "lightgray", level = 0.89) +
  facet_grid(Sex ~ life_history, 
             labeller = labeller(
               Sex = c("M" = "Male", "F" = "Female"),
               life_history = c("predispersal" = "Predispersal",
                                "floater" = "Floater",
                                "territory_holder" = "Reproductive adult")
             )) +
  theme_minimal() +
  labs(x = "Proportion of hours below 20ºC in past year", y = "Mass\n(g)") +
  theme(text = element_text(family = "Helvetica", size = 14),
        axis.text = element_text(size = 14),
        axis.text.y = element_text(size = 12, angle = 0, vjust = 0.5),
        axis.text.x = element_text(size = 6),
        axis.title.y = element_text(angle = 0, vjust = 0.5, size = 14),
        panel.spacing = unit(1, "lines"),
        strip.text.x = element_text(size = 9), 
        strip.text.y = element_text(size = 12, angle = 0, hjust = 0)) + 
  guides(color = FALSE, shape = FALSE) 
```

## Figure S5C
```{r figures5c, cache=TRUE}
ggplot(filter(biometrics7, !is.na(life_history) & !is.na(Sex)), aes(x = temp_utprop_year, y = Weight)) + 
  geom_point(aes(color = life_history, shape = life_history), size = 4, alpha = 0.7) +
  scale_color_manual(values = lhcolor_palette) +
  geom_smooth(method = "lm", color = "black", se = TRUE, fill = "lightgray", level = 0.89) +
  facet_grid(Sex ~ life_history, 
             labeller = labeller(
               Sex = c("M" = "Male", "F" = "Female"),
               life_history = c("predispersal" = "Predispersal",
                                "floater" = "Floater",
                                "territory_holder" = "Reproductive adult")
             )) +
  theme_minimal() +
  labs(x = "Proportion of hours above 35ºC in past year", y = "Mass\n(g)") +
  theme(text = element_text(family = "Helvetica", size = 14),
        axis.text = element_text(size = 14),
        axis.text.y = element_text(size = 12, angle = 0, vjust = 0.5),
        axis.text.x = element_text(size = 6),
        axis.title.y = element_text(angle = 0, vjust = 0.5, size = 14),
        panel.spacing = unit(1, "lines"),
        strip.text.x = element_text(size = 9),
        strip.text.y = element_text(size = 12, angle = 0, hjust = 0)) + 
  guides(color = FALSE, shape = FALSE) 
```

## Figure S6
```{r figures6, cache=TRUE}
set.seed(13)
# plot with dynamically calculated summary statistics
ggplot(filter(biometrics7, life_history == "territory_holder"), aes(x = Sex, y = Weight)) + 
  # add jittered points for each observation
  geom_jitter(size = 3, alpha = 0.5, aes(color = Sex), height = 0, width = 0.05, shape = "square") + 
  # add calculated mean points in black
  stat_summary(
    fun = mean, geom = "point", color = "black", # explicitly set color to black
    size = 6, shape = 18, position = position_dodge(width = 0.5)
  ) + 
  # add straight line for SD from the center point, with thicker line
  stat_summary(
    fun.data = function(x) {
      # calculate mean and SD manually
      mean_value <- mean(x, na.rm = TRUE)
      sd_value <- sd(x, na.rm = TRUE)
      # return the y-values for the SD line (mean +- 1 SD)
      return(data.frame(y = mean_value + c(-sd_value, sd_value), 
                        ymin = mean_value - sd_value, ymax = mean_value + sd_value))
    }, 
    geom = "errorbar", color = "black", # explicitly set color to black for errorbars
    size = 1, width = 0, position = position_dodge(width = 0.5) # increase line thickness with size = 2
  ) + 
  scale_color_manual(values = sexcolor_palette) + 
  theme_minimal() + 
  theme(
    text = element_text(family = "Helvetica", size = 15),
    axis.text = element_text(size = 14),
    axis.text.y = element_text(angle = 0, vjust = 0.5),
    axis.title.y = element_text(angle = 0, vjust = 0.5)
  ) + 
  labs(y = "Body\nmass\n(g)", x = "Sex") + 
  guides(color = FALSE)
```

## Figure S7
```{r figures7, cache=TRUE}
ggplot(filter(biometrics7, !is.na(life_history) & !is.na(Sex)), aes(x = Date, y = Bodylength)) +
  geom_point(aes(color = life_history, shape = life_history), size = 4, alpha = 0.7) +
  scale_color_manual(values = lhcolor_palette) +
  geom_smooth(method = "lm", fill = "lightgray", color = "black", level = 0.89) +
  facet_grid(Sex ~ life_history, 
             labeller = labeller(
               Sex = c("M" = "Male", "F" = "Female"),
               life_history = c("predispersal" = "Predispersal",
                                "floater" = "Floater",
                                "territory_holder" = "Reproductive adult")
             )) +
  theme_minimal() +
  guides(shape = FALSE, color = FALSE) +
  theme(text = element_text(family = "Helvetica", size = 14),
        axis.text = element_text(size = 14),
        axis.text.y = element_text(angle = 0, vjust = 0.5, size = 12),
        axis.text.x = element_text(size = 9),
        axis.title.y = element_text(angle = 0, vjust = 0.5, size = 14),
        panel.spacing = unit(1, "lines"),
        strip.text.x = element_text(size = 12),   # Life history label styling
        strip.text.y = element_text(size = 12, angle = 0, hjust = 0)) +  # Sex label styling
  labs(y = "Body\nlength\n(cm)", x = "Year") +
  scale_x_date(limits = as.Date(c("1999-01-01", "2023-12-31")), 
               breaks = seq(as.Date("1999-01-01"), as.Date("2023-12-31"), by = "6 years"),
               date_labels = "%Y")
```

## Figure S8
```{r figures8, cache=TRUE}
ggplot(filter(biometrics7, 
              life_history %in% c("territory_holder", 
                                  "floater", "predispersal")), 
       aes(x = day_of_year, y = Weight, color = life_history)) +
  geom_point(alpha = 0.5, size = 4, aes(shape = life_history)) +
  geom_smooth(color = "black",
              method = "gam", formula = y ~ s(x, bs = "cc"), 
              level = 0.89,
              method.args = list(family = "gaussian"), se = T, fill = "lightgray",
              linewidth = 1) +  
  scale_color_manual(values = lhcolor_palette) +
  theme_minimal() +
  labs(x = "Day of Year", y = "Mass (g)", 
       color = "Life History Stage", shape = "Life History Stage") +
    theme(text = element_text(family = "Helvetica", size = 15),
        axis.text = element_text(size = 14),
        axis.text.y = element_text(angle = 0, vjust = 0.5),
        axis.title.y = element_text(angle = 0, vjust = 0.5)) +
  facet_grid(Sex ~ life_history, 
             labeller = labeller(
               Sex = c("M" = "Male", "F" = "Female"),
               life_history = c("predispersal" = "Predispersal",
                                "floater" = "Floater",
                                "territory_holder" = "Reproductive adult")
             )) +
  guides(color = F, shape = F) +
  theme(text = element_text(family = "Helvetica", size = 14),
        axis.text = element_text(size = 14),
        axis.text.y = element_text(size = 12, angle = 0, vjust = 0.5),
        axis.text.x = element_text(size = 12),
        axis.title.y = element_text(angle = 0, vjust = 0.5, size = 14),
        panel.spacing = unit(1, "lines"),
        strip.text.x = element_text(size = 12),  
        strip.text.y = element_text(size = 12, angle = 0, hjust = 0)) +  
  guides(color = FALSE, shape = FALSE) 

```

## Figure S9a
```{r figures9a, cache=TRUE}
set.seed(13)
ggplot(filter(biometrics7, 
              BirthWithinYear != max(BirthWithinYear) &
                                       life_history == "territory_holder"),
       aes(x = Weight, y = BirthWithinYear)) +
  geom_jitter(size = 4, alpha = 0.5, aes(color = Sex), shape = "square",
              width = 0, height = 0.05) + 
  scale_color_manual(values = sexcolor_palette) +
  geom_smooth(method = "glm", method.args = list(family = "binomial"), 
              aes(color = Sex), fill = "lightgray", linewidth = 1.5) + 
  facet_wrap(~Sex, labeller = as_labeller(c("F" = "Female", "M" = "Male"))) + 
  theme_minimal() +
  scale_x_continuous(breaks = c(1100, 1300, 1500),
                      labels = c("1100", "1300", "1500")) +
  theme(
    text = element_text(family = "Helvetica", size = 15),
    axis.text = element_text(size = 14),
    axis.text.x = element_text(size = 11),
    axis.text.y = element_text(angle = 0, vjust = 0.5),
    axis.title.y = element_text(angle = 0, vjust = 0.5, size = 11),
    panel.spacing = unit(1.2, "lines")
  ) +
  labs(y = "Was there a  \nbirth in the group  \nin the last year?   ",
       x = "Body mass (g)") +
  guides(color = F)

```

## Figure S9b
```{r figures9b, cache=TRUE}
ggplot(filter(biometrics7, 
              BirthWithinNextYear != max(BirthWithinNextYear)  &
                                       life_history == "territory_holder"),
       aes(x = Weight, y = BirthWithinNextYear)) +
  geom_jitter(size = 4, alpha = 0.5, aes(color = Sex), shape = "square",
              width = 0, height = 0.05) + 
  scale_color_manual(values = sexcolor_palette) +
  geom_smooth(method = "glm", method.args = list(family = "binomial"), 
              aes(color = Sex), fill = "lightgray", linewidth = 1.5,
              level = 0.89) + 
  facet_wrap(~Sex, labeller = as_labeller(c("F" = "Female", "M" = "Male"))) + 
  scale_x_continuous(breaks = c(1100, 1300, 1500),
                      labels = c("1100", "1300", "1500")) +
  theme_minimal() +
  theme(
    text = element_text(family = "Helvetica", size = 15),
    axis.text = element_text(size = 14),
    axis.text.x = element_text(size = 11),
    axis.text.y = element_text(angle = 0, vjust = 0.5),
    axis.title.y = element_text(angle = 0, vjust = 0.5, size = 11),
    panel.spacing = unit(1.5, "lines")
  ) +
  labs(y = "Was there a   \nbirth in the group   \nin the next year?   ",
       x = "Body mass (g)") +
  guides(color = F)
```

## Figure S10
```{r figures10, cache=TRUE}
set.seed(13)
ggplot(filter(biometrics7, BirthWithinYear != max(BirthWithinYear) &
                                       life_history == "territory_holder"),
       aes(x = Date, y = BirthWithinYear)) +
  geom_jitter(size = 3, alpha = 0.7, height = 0.075, width = 0, 
              shape = "square", aes(color = Sex)) +
  scale_color_manual(values = sexcolor_palette, labels = c("F" = "Female", "M" = "Male")) +
  geom_smooth(aes(y = BirthWithinYear), method = "glm", 
              method.args = list(family = "binomial"), se = T,
              color = "black", fill = "lightgray",
              level = 0.89) + 
  theme_minimal() +
  theme(
    text = element_text(family = "Helvetica", size = 15),
    axis.text = element_text(size = 14),
    axis.text.y = element_text(angle = 0, vjust = 0.5),
     axis.title.y = element_text(angle = 0, vjust = 0.5, size = 12),
  ) +
  labs(y = "Was there a  \nbirth in the group  \nin the last year?   ",
       x = "Year", color = "Sex") +  # Add color legend title
  scale_x_date(limits = as.Date(c("1999-01-01", "2023-12-31")), 
               breaks = seq(as.Date("1999-01-01"), as.Date("2023-12-31"), by = "6 years"),
               date_labels = "%Y")
```

## Figure S11
```{r figures11, cache=TRUE}
#first scale the temperature data
numeric_columnsdaily <- sapply(daily, is.numeric)
daily_scaled <- daily
daily_scaled[, numeric_columnsdaily] <- scale(daily_scaled[, numeric_columnsdaily])
daily_scaled <- daily_scaled |> mutate(day_of_year = daily$day_of_year) |>
  mutate(decades_since_first_entry = as.numeric(difftime(Date, min(Date), units = "days")) / 3652.5)

ggplot(filter(biometrics7_scaled_th, BirthWithinYearfactor != "unknown"), aes(x = day_of_year)) + 
  geom_rect(xmin = 274, xmax = 334, ymin = -Inf, ymax = Inf, 
            fill = "lightblue", alpha = 0.2) +
  geom_text(x = 304, y = 1.5, label = "Birth\nSeason", 
            hjust = 0.5, color = "black", size = 4, family = "Helvetica") +
  geom_point(aes(y = Weight, color = life_history), size = 4, alpha = 0.7, 
             shape = "square") +
  scale_color_manual(values = lhcolor_palette) +
  geom_smooth(method = "gam", formula = y ~ s(x, bs = "cc"), 
              method.args = list(family = "gaussian"), 
              size = 2, aes(y = Weight, color = life_history),
              fill = "lightgray") +
  geom_smooth(data = daily_scaled, aes(x = day_of_year, y = mean_temp), 
              color = "maroon", method = "gam", formula = y ~ s(x, bs = "cc"), 
              size = 2, linetype = 2, se = FALSE, 
              method.args = list(family = "gaussian")) +
  theme_minimal() +
  labs(
    x = "Day of Year",
    shape = "Was there a\nbirth in the group\nin the last year?",
    y = "Scaled\nmean\ndaily\ntemperature\n(red)\n\nScaled\nbody\nmass\n(blue)"
  ) +
  facet_grid(
    Sex ~ BirthWithinYearfactor,
    labeller = labeller(
      Sex = c("M" = "Male", "F" = "Female"),
      BirthWithinYearfactor = c("no birth" = "No birth", "birth" = "Birth")
    )
  ) +
  theme(
    text = element_text(family = "Helvetica", size = 14),
    axis.text = element_text(size = 14),
    axis.text.y = element_text(size = 14, angle = 0, vjust = 0.5),
    axis.text.x = element_text(size = 12),
    axis.title.y = element_text(angle = 0, vjust = 0.5, size = 10),
    panel.spacing = unit(1, "lines"),
    strip.text.x = element_text(size = 12), 
    strip.text.y = element_text(size = 12, angle = 0, hjust = 0) 
  ) +
  guides(color = FALSE, shape = FALSE) +
  theme(
  panel.background = element_rect(fill = "white", color = NA),
  plot.background = element_rect(fill = "white", color = NA)
)
```