---
title: "Monkey_Mass"
author: "Jonathan Pertile"
date: "2024-12-09"
output: html_document
---

This includes the packages used in the code.
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyr)
library(ggplot2)
library(lubridate)
library(brms)
library(ggdag)
library(pals)
library(cetcolor)
library(lme4)
library(ggtext)
library(parallel)
library(truncnorm)
library(purrr)
library(MASS)
library(zoo)
```

# Temperature

## Temperature Data
This is the hourly temperature data from El Pucú Airport in Formosa (hourly). 
It has also been summarized at the day-level scale (daily).
```{r}
# load hourly temperature data and separate date and time components
hourly <- read.csv("./hourlytemp.csv") |>
  separate(DATE, into = c("Date", "Time"), sep = "T")

# convert date column to date format
hourly$Date <- as.Date(hourly$Date)

# filter to retain only hourly records at the start of each hour
hourly <- hourly |>
  filter(substr(Time, 4, 5) == "00" & substr(Time, 7, 8) == "00")

# aggregate hourly data into daily summaries
daily <- hourly %>%
  group_by(Date) %>%
  filter(sum(!is.na(temp)) / n() >= 0.95) %>%  # retain days with at least 95% valid data
  summarize(
    mean_temp = mean(temp, na.rm = TRUE),  # calculate mean temperature
    prop_above_35 = mean(temp > 35, na.rm = TRUE),  # proportion of hourly temperatures above 35
    prop_below_20 = mean(temp < 20, na.rm = TRUE)  # proportion of hourly temperatures below 20
  )

# add a column for month and day
daily <- mutate(daily, month_day = format(Date, "%m-%d"))

# calculate the day of the year from month and day
daily$day_of_year <- as.numeric(format(as.Date(daily$month_day, format = "%m-%d"), "%j"))

# process hourly data with similar columns
hourly <- hourly %>%
  mutate(
    month_day = format(Date, "%m-%d"),
    day_of_year = as.numeric(format(as.Date(month_day, format = "%m-%d"), "%j"))
  )

```

This is the temperature data from Estancia Guaycolec (temperature_eg). 
It has also been summarized at the day-level scale (temperature_eg_daily).
```{r}
# read temperature data and convert date column
temperature_eg <- read.csv("./Temp-edited2_updated02Dec23v3.csv")
temperature_eg$Date <- as.Date(temperature_eg$Date)

# create daily summaries for temperature data
temperature_eg_daily <- temperature_eg %>%
  group_by(Date) %>%
  filter(sum(!is.na(Temperature)) / n() >= 0.95) %>%  # retain days with at least 95% valid data
  summarize(
    mean_temp = mean(Temperature, na.rm = TRUE),  # mean temperature
    prop_above_35 = mean(Temperature > 35, na.rm = TRUE),  # proportion above 35
    prop_below_20 = mean(Temperature < 20, na.rm = TRUE),  # proportion below 20
    variance = var(Temperature, na.rm = TRUE)  # variance of temperature
  )

# add columns for month-day and day of the year to daily data
temperature_eg_daily <- temperature_eg_daily %>%
  mutate(
    month_day = format(Date, "%m-%d"),
    day_of_year = as.numeric(format(as.Date(month_day, format = "%m-%d"), "%j"))
  )

# repeat similar processing for the full dataset
temperature_eg <- temperature_eg %>%
  mutate(
    month_day = format(Date, "%m-%d"),
    day_of_year = as.numeric(format(as.Date(month_day, format = "%m-%d"), "%j"))
  )
```

## Temperature Functions
These functions calculate temperature values in the past month. 
```{r}
last_month_temp <- function(Date) {
  # convert input to a date object
  input_date <- as.Date(Date)
  
  # create a range of dates covering the last 30 days
  date_range <- input_date - 30:0
  
  # subset the data for the specified date range
  subset_data <- filter(daily, Date %in% date_range)
  
  # check if at least 95% of the data is available
  if (length(subset_data$mean_temp) / 31 >= 0.95) {
    # calculate the average temperature
    avg_temp <- mean(subset_data$mean_temp, na.rm = TRUE)
    return(avg_temp)
  } else {
    return(NA)  # return NA if insufficient data
  }
}

last_month_temp_ut_prop <- function(Date) {
  # convert input to a date object
  input_date <- as.Date(Date)
  
  # create a range of dates covering the last 30 days
  date_range <- input_date - 30:0
  
  # subset the data for the specified date range
  subset_data <- filter(daily, Date %in% date_range)
  
  # check if at least 95% of the data is available
  if (length(subset_data$prop_above_35) / 31 >= 0.95) {
    # calculate the mean proportion above 35
    mean_above_35 <- mean(subset_data$prop_above_35, na.rm = TRUE)
    return(mean_above_35)
  } else {
    return(NA)  # return NA if insufficient data
  }
}
```

These functions calculate temperature values in the past year. 
```{r}
last_year_temp <- function(Date) {
  # convert input to a date object
  input_date <- as.Date(Date)
  
  # create a range of dates covering the last 365 days
  date_range <- input_date - 365:0
  
  # subset the data for the specified date range
  subset_data <- filter(daily, Date %in% date_range)
  
  # check if at least 95% of the data is available
  if (length(subset_data$mean_temp) / 365 >= 0.95) {
    # calculate the average temperature
    avg_temp <- mean(subset_data$mean_temp, na.rm = TRUE)
    return(avg_temp)
  } else {
    return(NA)  # return NA if insufficient data
  }
}

last_year_temp_ut_prop <- function(Date) {
  # convert input to a date object
  input_date <- as.Date(Date)
  
  # create a range of dates covering the last 365 days
  date_range <- input_date - 365:0
  
  # subset the data for the specified date range
  subset_data <- filter(daily, Date %in% date_range)
  
  # check if at least 95% of the data is available
  if (length(subset_data$prop_above_35) / 365 >= 0.95) {
    # calculate the mean proportion above 35
    mean_above_35 <- mean(subset_data$prop_above_35, na.rm = TRUE)
    return(mean_above_35)
  } else {
    return(NA)  # return NA if insufficient data
  }
}

last_year_temp_lt_prop <- function(Date) {
  # convert input to a date object
  input_date <- as.Date(Date)
  
  # create a range of dates covering the last 365 days
  date_range <- input_date - 365:0
  
  # subset the data for the specified date range
  subset_data <- filter(daily, Date %in% date_range)
  
  # check if at least 95% of the data is available
  if (length(subset_data$prop_below_20) / 365 >= 0.95) {
    # calculate the mean proportion below 20
    mean_below_20 <- mean(subset_data$prop_below_20, na.rm = TRUE)
    return(mean_below_20)
  } else {
    return(NA)  # return NA if insufficient data
  }
}

```


# Dataset

```{r}
#load data
biometrics7 <- read.csv("./biometrics7.csv")

biometrics7$Sex <- as.factor(biometrics7$Sex)  # convert sex to factor
sexcolor_palette <- c("F" = "#498467", "M" = "#9067c6")  # color palette for sex

biometrics7$life_history <- factor(biometrics7$life_history, levels = c("predispersal", "floater", "territory_holder"))  # reorder life_history factor
lhcolor_palette <- c("predispersal" = "#DF6873", "floater" = "#AA5EE8", "territory_holder" = "#9DACFF")  # color palette for life history

# create ordered factors for birth status variables
biometrics7$BirthWithinYearfactor <- factor(biometrics7$BirthWithinYear, 
                                            levels = c(0, 2, 1), 
                                            labels = c("no birth", "unknown", "birth"), 
                                            ordered = TRUE)
biometrics7$BirthWithinNextYearfactor <- factor(biometrics7$BirthWithinNextYear, 
                                                levels = c(0, 2, 1), 
                                                labels = c("no birth", "unknown", "birth"), 
                                                ordered = TRUE)
biometrics7$BirthWithinLast5Monthsfactor <- factor(biometrics7$BirthWithinLast5Months, 
                                                   levels = c(0, 2, 1), 
                                                   labels = c("no birth", "unknown", "birth"), 
                                                   ordered = TRUE)

# standardize biometric data, and then unscale the variables I do not want to be scaled
numeric_columns <- sapply(biometrics7, is.numeric)
biometrics7_scaled <- biometrics7 |> 
  mutate(across(all_of(names(numeric_columns)[numeric_columns]), scale)) |> 
  mutate(day_of_year = biometrics7$day_of_year, 
         time_since_first_entry = biometrics7$time_since_first_entry,
         decades_since_first_entry = biometrics7$decades_since_first_entry,
         BirthWithinYear = biometrics7$BirthWithinYear,
         BirthWithinNextYear = biometrics7$BirthWithinNextYear,
         BirthWithinLast5Months = biometrics7$BirthWithinLast5Months,
         age = biometrics7$age)

# scale territory_holder data,
#and then unscale the variables I do not want to be scaled
biometrics7_scaled_th <- filter(biometrics7, life_history == "territory_holder")
biometrics7_scaled_th[, numeric_columns] <- scale(biometrics7_scaled_th[, numeric_columns])
biometrics7_scaled_th <- biometrics7_scaled_th |> mutate(day_of_year = filter(biometrics7, life_history == "territory_holder")$day_of_year) |>
  mutate(time_since_first_entry = filter(biometrics7, life_history == "territory_holder")$time_since_first_entry) |>
  mutate(decades_since_first_entry = filter(biometrics7, life_history == "territory_holder")$decades_since_first_entry) |>
  mutate(BirthWithinYear = filter(biometrics7, life_history == "territory_holder")$BirthWithinYear) |>
  mutate(BirthWithinNextYear = filter(biometrics7, life_history == "territory_holder")$BirthWithinNextYear) |>
  mutate(BirthWithinLast5Months = filter(biometrics7, life_history == "territory_holder")$BirthWithinLast5Months)


```
